A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-29 paper.tex
minor

We extend the bounded synthesis algorithm to learn states losing for one of the players from
failed attempts to find candidate strategies.  The learning procedure kicks in whenever
\textsc{findCandidate} cannot find a candidate strategy for an abstract game tree with initial
state $s$.  We know that $s$ is a losing state.  We can learn additional losing
states from the tree via interpolation.  This is achieved in lines~18--20 in Algorithm~\ref{alg:bounded},
enabled in the unbounded version of the algorithm, which invoke \textsc{learn} or
\textsc{\textoverline{learn}} to learn controller or environment losing states respectively
(Algorithm~\ref{alg:learn}).

@@

Consider the game tree $T$ in
Figure~\ref{fig:interpolatetree}, whose edges are labelled with environment
actions.  Assume that there does not exist a partial strategy for the controller
winning against this tree and that the \textsc{findCandidate} function could not
find a candidate strategy.
We choose a non-leaf node $n$ of $T$ with maximal depth, i.e., a node whose children
are leafs (Algorithm~\ref{alg:learn}, line~3). We then split the tree at $n$ such that both slices $T_1$ and $T_2$
contain a copy of $n$ (line~4). Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$
except $n$'s children, and $T_2$ (Figure~\ref{fig:treef2}), which contains only $n$ and its children.  
There is no candidate strategy for $T$ so $s \land
\textsc{treeFormula}(T)$ is unsatisfiable.  By construction, 
$\textsc{treeFormula}(T) = \textsc{treeFormula}(T_1) \land \textsc{treeFormula}(T_2)$
and hence $s \land
\textsc{treeFormula}(T_1) \land \textsc{treeFormula}(T_2)$ is also 
unsatisfiable. This enables the construction of an \emph{interpolant} that
captures losing states at node $n$.

@@

We can construct an interpolant with $F_1 = s \land \textsc{treeFormula}(T_1)$
and $F_2 = \textsc{treeFormula}(T_2)$ (line~5). The only variables shared between $F_1$
and $F_2$ are the state variable copies belonging to node $n$. By construction
$F_2 \land \mathcal{I}$ is unsatisfiable, therefore all states in $\mathcal{I}$
are losing against abstract game tree $T_2$ in Figure~\ref{fig:treef2}.  We also know that $F_1 \to
\mathcal{I}$, thus $\mathcal{I}$ contains all states reachable at $n$ by
following $T_1$ and avoiding error states.

@@

Learning of states losing for the environment is similar (\textsc{\textoverline{learn}} in Algorithm~\ref{alg:learn}). 
The main difference is that environment-losing states are only losing for a particular
bound: given that there does not exist an environment strategy that
forces the game into an error state in $k$ rounds or less; there may still exist a longer
environment-winning strategy.  We therefore record learned environment-losing states
along with associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.  $B^m[k]$ 
are initialized to $\top$ for all $k$.  Whenever an environment-losing set $\mathcal{I}$
is discovered for a node $n$ with bound $\textsc{bound}(n)$ in line~13, Algorithm~\ref{alg:learn}, this set is subtracted from 
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).

@@

@@

The sets $B^m$ and $B^M$ are only modified by the functions \textsc{learn} and
\textoverline{\textsc{learn}}.  Below we prove that these functions indeed
maintain the invariants.

\subsection{Proof of learning functions}

>>>>

We extend the bounded synthesis algorithm to learn states losing for one of the players from
failed attempts to find candidate strategies.  The learning procedure kicks in whenever
\textsc{findCandidate} cannot find a candidate strategy for an abstract game tree with initial
state $s$.  At this point we have established that $s$ is a losing state.  We can learn additional losing
states from the tree via interpolation.  This is achieved in lines~18--20 in Algorithm~\ref{alg:bounded},
enabled in the unbounded version of the algorithm, which invoke \textsc{learn} or
\textsc{\textoverline{learn}} to learn controller or environment losing states respectively
(Algorithm~\ref{alg:learn}).

@@

Consider the game tree $T$ in
Figure~\ref{fig:interpolatetree}, whose edges are labelled with environment
actions.  Assume that there does not exist a partial strategy for the controller
winning against this tree and that the \textsc{findCandidate} function could not
find a candidate strategy.
We choose a non-leaf node $n$ of $T$ with maximal depth, i.e., a node whose children
are leafs (Algorithm~\ref{alg:learn}, line~3). We then split the tree at $n$ such that both slices $T_1$ and $T_2$
contain a copy of $n$ (line~4). Figure~\ref{fig:treef1} shows $T_1$, which contains all of $T$
except $n$'s children, and $T_2$ (Figure~\ref{fig:treef2}), which contains only $n$ and its children.  
There is no candidate strategy for $T$ so $s \land
\textsc{treeFormula}(T)$ is unsatisfiable.  By construction, 
$\textsc{treeFormula}(T) \equiv \textsc{treeFormula}(T_1) \land \textsc{treeFormula}(T_2)$
and hence $s \land
\textsc{treeFormula}(T_1) \land \textsc{treeFormula}(T_2)$ is also 
unsatisfiable. This enables the construction of an \emph{interpolant} that
captures losing states at node $n$.

@@

We construct an interpolant with $F_1 = s(X_T) \land \textsc{treeFormula}(T_1)$
and $F_2 = \textsc{treeFormula}(T_2)$ (line~5). The only variables shared between $F_1$
and $F_2$ are the state variable copies belonging to node $n$. By the properties of the interpolant,
$F_2 \land \mathcal{I}$ is unsatisfiable, therefore all states in $\mathcal{I}$
are losing against abstract game tree $T_2$ in Figure~\ref{fig:treef2}.  We also know that $F_1 \to
\mathcal{I}$, thus $\mathcal{I}$ contains all states reachable at $n$ by
following $T_1$ and avoiding error states.

@@

Learning of states losing for the environment is similar (\textsc{\textoverline{learn}} in Algorithm~\ref{alg:learn}). 
The main difference is that environment-losing states are only losing for a particular
bound: given that there does not exist an environment strategy that
forces the game into an error state in $k$ rounds or less; there may still exist a longer
environment-winning strategy.  We therefore record learned environment-losing states
along with associated bounds.  To this end, we maintain a conceptually infinite array of
sets $B^m[k]$ that are may-losing for the controller, indexed by bound $k$.  $B^m[k]$ 
are initialized to $\top$ for all $k$.  Whenever an environment-losing set $\mathcal{I}$
is discovered for a node $n$ with bound $\textsc{bound}(n)$ in line~13 of Algorithm~\ref{alg:learn}, this set is subtracted from 
$B^m[i]$, for all $i$ less than or equal to the bound (lines~14--16).

@@

\subsection{Main synthesis loop}

@@

Both invariants trivially hold after $B^m$ and $B^M$ have been initialized in the
beginning of the algorithm.
The sets $B^m$ and $B^M$ are only modified by the functions \textsc{learn} and
\textoverline{\textsc{learn}}.  Below we prove that \textoverline{\textsc{learn}} 
maintains the invariants.  The proof of \textsc{learn} is similar.

\subsection{Proof of \textoverline{\textsc{learn}}}
